Multi-Agent Pathfinding with Matching using Increasing Cost Tree Search

### 2021-06-22 docs/paper.tex
Corrected some errors mentioned in feedback

	Little to no research has been done on MAPFM and solving it optimally. For MAPFM with a makespan objective instead of SIC, known as the combined target-assignment and pathfinding (TAPF) problem, there is the Conflict-Based Min-Cost-Flow (CBM) algorithm by Ma and Koenig \cite{ma2016} and a similar approach described in \cite{henkel2019}. Ma and Koenig's method builds upon the conflict-based search method and in particular on the Meta-Agent variation thereof also discussed in \cite{sharon2015} as a high-level search framework; in the low-level search, the connection between anonymous MAPF problems and max-flow problems (as discussed in \cite{yu2013}) is exploited to plan single teams in polynomial time.

@@

	As discussed in \cite{ma2016}, MAPFM generalizes both MAPF and anonymous MAPF, which is MAPF allowing an agent to move to any goal, a problem that can be solved in polynomial time.
	\section{Solving MAPF with Matching using ICTS} % Your contribution
	\label{section:icts-matching}
	In this section, taking ICTS for MAPF as a starting point, two methods to use ICTS for MAPFM are described. The first method relies on a reduction from MAPFM to repeated MAPF: an optimal solution to MAPFM corresponds to a matching of agents to goals, so by exhaustively enumerating all matchings and solving these as MAPF instances, an optimal solution to the MAPFM instance can be found. In the second method, the ICT search itself is modified to allow an agent $a_j^i$ to be matched to any $g_k^i$.
	\subsection{Increasing Cost Tree Search}
	\label{icts}
	In Increasing Cost Tree Search\cite{sharon2011}, the two-step mapping from path combination to SIC described in Section \ref{mapfm} is reversed to search for a solution with minimal SIC. This is done by searching for cost vectors corresponding to increasing SIC on the top level and searching for non-conflicting path combinations (solutions) for each cost vector on the bottom level.

	\subsubsection{Top-level search}
	In the top-level search, all possible cost vectors for a given SIC are evaluated by searching an Increasing Cost Tree as depicted in Figure \ref{fig:ict}, starting from $C^* = \sum_i c^*_i$ corresponding to a single root cost vector $(c^*_1,c^*_2,\ldots,c^*_k)$. This root has $k$ children $(c^*_1 + 1,c^*_2,\ldots,c^*_k),(c^*_1,c^*_2 + 1,\ldots,c^*_k),\ldots,(c^*_1,c^*_2,\ldots,c^*_k + 1)$ all of cost $C = C^* + 1$. Searching this ICT breadth-first corresponds to evaluating all possible cost vectors corresponding to increasing cost $C$ starting with $C^*$, guaranteeing optimality of a found solution. The number of cost vectors evaluated when searching up to nodes of cost $C^* + \Delta$ is $\mathcal{O}(k^\Delta)$. Node evaluation is done by a low-level search that searches for a solution $(\pi_1,\ldots,\pi_k)$ corresponding to the node costs $(c_1,\ldots,c_k)$.

@@

			\EndProcedure
		\end{algorithmic}
		\caption{Increasing Cost Tree Search}
		\label{icts-algo}
	\end{algorithm}
	\subsection{Exhaustive ICTS}
	\label{exhaustive}
	For $K$ teams $t_1,\ldots,t_K$ of size $k_i$, there are in total $M = \prod_{i} k_i!$ matchings from team agents to team goals. Let $C_1$ through $C_M$ denote the optimal SIC costs of each MAPF instance corresponding to the MAPFM instance, with $C_m = \infty$ for infeasible $m$. In order to find the optimal solution, $m$ with lowest $C_m$ has to be found, which requires all $M$ to be solved or found to be unfeasible, meaning that $\mathcal{O}(M)$ MAPF instances have to be solved. 
	\subsubsection{Bounded search}
	Once a solution is found for a matching $m$ with cost $C_m$, this can be used in the next search as an (exclusive) upper-bound $B$ for the SIC of nodes that have to be searched. Formally, let $B_0 = \infty$ represent the initial SIC bound of the ICT search. After solving each $m$ with solution cost $C_m$, let $B_{m+1} = \min(B_m,C_m)$. By successively updating this $B$ in this manner, as more matchings are solved, the effort needed to solve the remaining matchings can be expected to be reduced significantly. With $C_m$ denoting the optimal SIC for matching $m$, this effort can be expressed as	
	\[\Delta_{m,\text{bounded}} = \min(B_m,C_m) - C^*_m = B_{m+1} - C^*_m\]
	This is proportional to the effort due to the complexity of the ICTS search for matching $m$, $\mathcal{O}(k^{\Delta_{m,\text{bounded}}})$.
	If $B_m < C^*_m$, only the root node will be generated to determine this so $\Delta_{m,\text{bounded}} = 0$.
	Using this, the number of ICT nodes searched in total becomes $\mathcal{O}(\sum_m k^{\Delta_{m,\text{bounded}}})$.
	\subsubsection{Ordered enumeration}
	\label{ordered-enum}
	To reduce the total effort, the matchings $m$ can be enumerated and solved in increasing order of $C^*_m$. In this case, the effort per matching is monotonically non-increasing by Property \ref{monotonicity}. This induces a case in which this enumeration minimizes the number of nodes searched in total (Property \ref{cond-monoton-optim}). However, cases in which $\Delta_m$ of the first $m$'s searched is much higher could also be constructed, in which case this approach might cause more nodes to be processed in total than alternative methods. Therefore, this ordered enumeration remains a heuristic improvement that works well in practice, as shown in Section \ref{experiments}.

	\begin{property}[Monotonicity]
		\label{monotonicity}
		In bounded exhaustive search using ICTS, solving the matchings in increasing order of root SIC makes the effort $\Delta_{m,\text{bounded}}$ monotonically non-increasing as a function of $m$.
	\end{property}
	\begin{proof}
		Due to the order of matchings for successive matchings $m,m+t$ for some $t\geq 1$, $C^*_{m} \leq C^*_{m+t}$. By definition, $B_{m+2} = \min(B_{m+1},C_{m+1})$ meaning that $B_{m+2} \leq B_{m + 1}$ so transitively $B_{m+t+1} \leq B_{m+1}$. Therefore
		\[B_{m+t+1} - C^*_{m+t}\leq B_{m+1} - C^*_m\implies \Delta_{m+t,\text{bounded}} \leq \Delta_{m,\text{bounded}}\]
	\end{proof}
	\begin{property}[Conditional optimality]
		\label{cond-monoton-optim}
		Assuming that $\Delta_m$, the unbounded effort, is monotonically non-decreasing as a function of $m$ (with $m$ in increasing order of $C^*_m$), ordered enumeration minimizes the total number of nodes searched across matchings.
	\end{property}
	\subsubsection{Approximate ordered enumeration}
	In practice, it is difficult enumerate all matchings in increasing order of root SIC. Finding a lowest SIC $m$ reduces to the assignment problem that can be solved using the Hungarian algorithm \cite{kuhn55} in polynomial time; however, to the knowledge of the author, no algorithm exists to more generally compute the $i$'th matching in increasing order of $C^*_m$ (assuming some fixed tie-breaking). This means that matchings have to be exhaustively enumerated and stored in memory to sort them, but this requires $\mathcal{O}(M)$ memory which was found to be more problematic in practice than the time complexity of the sorting, $\mathcal{O}(M\log(M))$.

@@

	\subsection{ICTS-m}
	ICTS with Matching (ICTS-m) generalizes ICTS to optimally solve MAPFM instances. Few changes to ICTS were necessary to accommodate for matching in this increasing cost tree scheme. Specifically, the ICT root generation and the MDD generation as described in \ref{icts} had to be changed.
	\subsubsection{Root generation for matching}
	Given a team $t_i$ of size $k_i$, for any agent $a_j^i$ there are $k_i$ costs $c_j(1),c_j(2),\ldots,c_j(k_i)$ for shortest paths from $s_j^i$ to each goal $g_k^i$. Making no assumptions about what goals other agents in $t_i$ are assigned to, it could be that in the optimal solution $a_j^i$ is assigned to a goal $g_{m}^i$ with path cost $c^*_j = c_j(m)$ such that $c_j(m) = \min_{n\in\{0,\ldots,k_i\}} c_j(n)$. Therefore, to guarantee optimality, this minimal cost to any matching goal has to take the place of the shortest path cost from the original ICTS root. Intuitively, the sum of individual costs of this root is highly optimistic, particularly if the teams are large. This is because often, agents will not move to their closest matching goal in an optimal solution, even without considering conflicts (see \ref{pathology}).

@@

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.40\textwidth}
			\centering
			\includegraphics[width=\linewidth]{img/counter-example-ep-2}
			\caption{Instance}
			\label{fig:counter-example:a}
		\end{subfigure}
		\begin{subfigure}{0.49\textwidth}
			\centering
			\includegraphics[width=\linewidth]{img/counter-example-tree}
			\caption{Pruned ICT that is traversed}
			\label{fig:counter-example:b}
		\end{subfigure}
		\caption{Counter-example for enhanced pruning combined with child pruning}
		\label{fig:counter-example}
	\end{figure}
	\begin{proof}
		Consider Figure \ref{fig:counter-example:a} with a red agent $r$, green agent $g$ and black agent $b$. It can be seen that $(3,8,5)$ with SIC $16$ is an optimal solution for this instance, which corresponds to $r$ moving aside, allowing $g$ to pass.  Say enhanced pruning with pairs is applied to these three agents at $(c^*_r,c^*_g,c^*_b) = (0,8,5)$. Let the enhanced pruning order be $(r,g)$, $(r,b)$, $(g,b)$ without loss of generality. $(r,g)$ is solvable but will leave for $g$ out of the two paths of length $8$ only the clock-wise path. $(r,b)$ will not change either MDD in this node. When $(g,b)$ is evaluated, it is found that the remaining clockwise path for $g$ and $b$'s path are incompatible (edge conflict) so $(0,9,5)$ and $(0,8,6)$ are generated. By repeating this order of enhanced pruning with child pruning, the ICT shown in \ref{fig:counter-example:b} is traversed. Notice that none of the nodes with SIC 16 in this ICT are a solution, meaning that enhanced pruning combined with child pruning will return a suboptimal solution to the instance in \ref{fig:counter-example:a}, making this pruning strategy suboptimal.
	\end{proof}

@@

	CBM solving all problems with one team (Figure \ref{fig:r-25-1-p} and \ref{fig:r-75-1-p}), problems equivalent to anonymous MAPF instances, is at first surprising but follows directly from the definition of the CBM method. CBM solves single teams as anonymous MAPF instances in polynomial time, meaning that solving a single team reduces to a polynomial-time procedure, namely SSP. Following the results reported in \cite{ma2016}, it is expected that as the number of teams approaches $k$, CBM performance will deteriorate and thus become more similar to the results for exhaustive methods here reported.
	%	Still, some implementations might be more optimized than others so this might be a factor in performance characteristics and in particular in relative performance differences.
	\section{Conclusions and Future work}
	\label{conclusions}
	\label{future}
	The goal of this research was to seek alternative approaches to solving MAPFM based on ICTS. In this, the research has succeeded, but the scalability of the algorithms leaves something to be desired. Exhaustive methods, like those discussed in detail in \ref{exhaustive}, cannot be expected to scale well as teams grow larger due to their complexity, no matter what optimizations are put in place. Nonetheless, as was shown in Section \ref{experiments}, the more advanced exhaustive ICTS algorithms consistently outperform the ICTS-m based algorithms. Compared to the SIC-adaptation of the CBM method \cite{baauw2021}, it was seen that exhaustive methods generally perform relatively poorly if there are fewer teams and fewer obstacles, while being competitive once teams grow smaller and maps become more constrained. Delineating under which conditions exhaustive methods outperform CBM and vice versa is something for further research.

>>>>

	Comparatively little research has been done on MAPFM and solving it optimally. For MAPFM with a makespan objective instead of SIC, known as the combined target-assignment and pathfinding (TAPF) problem, there is the Conflict-Based Min-Cost-Flow (CBM) algorithm by Ma and Koenig \cite{ma2016} and a similar approach described in \cite{henkel2019}. Ma and Koenig's method builds upon the conflict-based search method and in particular on the Meta-Agent variation thereof also discussed in \cite{sharon2015} as a high-level search framework; in the low-level search, the connection between anonymous MAPF problems and max-flow problems (as discussed in \cite{yu2013}) is exploited to plan single teams in polynomial time.

@@

	As discussed in \cite{ma2016}, MAPFM generalizes both MAPF and anonymous MAPF, which is MAPF allowing an agent to move to any goal, a problem that can be solved in polynomial time.
	\section{Solving MAPF with Matching using ICTS} % Your contribution
	\label{section:icts-matching}
	In this section, taking ICTS for MAPF as a starting point, two methods to use ICTS for MAPFM are described. The first method relies on a reduction from MAPFM to repeated MAPF: an optimal solution to MAPFM corresponds to a matching of agents to goals, so by exhaustively enumerating all matchings and solving these as MAPF instances, an optimal solution to the MAPFM instance can be found. In the second method, the ICT search itself is modified to allow an agent $a_j^i$ to be matched to any $g_k^i$.
	\subsection{Increasing Cost Tree Search}
	\label{icts}
	In Increasing Cost Tree Search\cite{sharon2011}, the two-step mapping from path combination to SIC described in Section \ref{mapfm} is reversed to search for a solution with minimal SIC. This is done by searching for cost vectors corresponding to increasing SIC at the top level and searching for non-conflicting path combinations (solutions) for each cost vector at the bottom level.

	\subsubsection{Top-level search}
	In the top-level search, all possible cost vectors for a given SIC are evaluated by searching an Increasing Cost Tree as depicted in Figure \ref{fig:ict}, starting from $C^* = \sum_i c^*_i$ corresponding to a single root cost vector $(c^*_1,c^*_2,\ldots,c^*_k)$. This root has $k$ children $(c^*_1 + 1,c^*_2,\ldots,c^*_k),(c^*_1,c^*_2 + 1,\ldots,c^*_k),\ldots,(c^*_1,c^*_2,\ldots,c^*_k + 1)$ all of cost $C = C^* + 1$. Searching this ICT breadth-first corresponds to evaluating all possible cost vectors corresponding to increasing cost $C$ starting with $C^*$, guaranteeing optimality of a found solution. When searching up to  cost $C^* + \Delta$, $\mathcal{O}(k^\Delta)$ cost vectors are evaluated. Node evaluation corresponds to performing a low-level search for a solution $(\pi_1,\ldots,\pi_k)$ corresponding to the node cost vector $(c_1,\ldots,c_k)$.

@@

			\EndProcedure
		\end{algorithmic}
		\caption{Increasing Cost Tree Search}
		\label{icts-algo}
	\end{algorithm}
	\subsection{Exhaustive ICTS}
	\label{exhaustive}
	For $K$ teams $t_1,\ldots,t_K$ of size $k_i$, there are in total $M = \prod_{i} k_i!$ matchings from team agents to team goals. Let $C_1$ through $C_M$ denote the optimal SIC costs of each MAPF instance corresponding to the MAPFM instance, with $C_m = \infty$ for infeasible $m$. In order to find the optimal solution, $m$ with lowest $C_m$ has to be found, which requires all $M$ to be solved or found to be infeasible, meaning that $\mathcal{O}(M)$ MAPF instances have to be solved. 
	\subsubsection{Bounded search}
	Once a solution is found for a matching $m$ with cost $C_m$, this can be used in the next search as an (exclusive) upper-bound $B$ for the SIC of nodes that have to be searched. Formally, let $B_0 = \infty$ represent the initial SIC bound of the ICT search. After solving each $m$ with solution cost $C_m$, let $B_{m+1} = \min(B_m,C_m)$. By successively updating this $B$ in this manner, as more matchings are solved, the effort needed to solve the remaining matchings can be expected to be reduced significantly. With $C_m$ denoting the optimal SIC for matching $m$, this effort can be expressed as	
	\[\Delta_{m,\text{bounded}} = \min(B_m,C_m) - C^*_m = B_{m+1} - C^*_m\]
	This is proportional to the effort due to the complexity of the ICTS search for matching $m$, $\mathcal{O}(k^{\Delta_{m,\text{bounded}}})$.
	If $B_m < C^*_m$, only the root node will be generated to determine this so $\Delta_{m,\text{bounded}} = 0$.
	Using this, the number of ICT nodes searched in total becomes $\mathcal{O}(\sum_m k^{\Delta_{m,\text{bounded}}})$.
	\subsubsection{Ordered enumeration}
	\label{ordered-enum}
	To reduce the total effort, the matchings $m$ can be enumerated and solved in increasing order of $C^*_m$. In this case, the effort per matching is monotonically non-increasing by Property \ref{monotonicity}. This induces a case in which this enumeration minimizes the number of nodes searched in total (Property \ref{cond-monoton-optim}). However, cases in which $\Delta_m$ of the first $m$'s searched is much higher could also be constructed, in which case this approach might cause more nodes to be processed in total than alternative methods. Therefore, this ordered enumeration remains a heuristic improvement that works well in practice, as shown in Section \ref{experiments}.

	\begin{property}[Monotonicity]
		\label{monotonicity}
		In bounded exhaustive search using ICTS, solving the matchings in increasing order of root SIC makes the effort $\Delta_{m,\text{bounded}}$ monotonically non-increasing as a function of $m$.
	\end{property}
	\begin{proof}
		Due to the order of matchings for successive matchings $m,m+t$ for some $t\geq 1$, $C^*_{m} \leq C^*_{m+t}$. By definition, $B_{m+2} = \min(B_{m+1},C_{m+1})$ meaning that $B_{m+2} \leq B_{m + 1}$ so transitively $B_{m+t+1} \leq B_{m+1}$. Therefore
		\[B_{m+t+1} - C^*_{m+t}\leq B_{m+1} - C^*_m\implies \Delta_{m+t,\text{bounded}} \leq \Delta_{m,\text{bounded}}\]
	\end{proof}
	\begin{property}[Conditional optimality]
		\label{cond-monoton-optim}
		Assuming that $\Delta_m$, the unbounded effort, is monotonically non-decreasing as a function of $m$ (with $m$ in increasing order of $C^*_m$), ordered enumeration minimizes the total number of nodes searched across matchings.
	\end{property}
	\subsubsection{Approximate ordered enumeration}
	In practice, it is difficult to enumerate all matchings in increasing order of root SIC. Finding a lowest SIC $m$ reduces to the assignment problem that can be solved using the Hungarian algorithm \cite{kuhn55} in polynomial time; however, to the knowledge of the author, no algorithm exists to more generally compute the $i$'th matching in increasing order of $C^*_m$ (assuming some fixed tie-breaking). This means that matchings have to be exhaustively enumerated and stored in memory to sort them, but this requires $\mathcal{O}(M)$ memory which was found to be more problematic in practice than the time complexity of the sorting, $\mathcal{O}(M\log(M))$.

@@

	\subsection{ICTS-m}
	ICTS with Matching (ICTS-m) generalizes ICTS to optimally solve MAPFM instances. Few changes to ICTS were necessary to accommodate for matching in this increasing cost tree scheme. Specifically, the ICT root generation and the MDD generation as described in Subsection \ref{icts} had to be changed.
	\subsubsection{Root generation for matching}
	Given a team $t_i$ of size $k_i$, for any agent $a_j^i$ there are $k_i$ costs $c_j(1),c_j(2),\ldots,c_j(k_i)$ for shortest paths from $s_j^i$ to each goal $g_k^i$. Making no assumptions about what goals other agents in $t_i$ are assigned to, it could be that in the optimal solution $a_j^i$ is assigned to a goal $g_{m}^i$ with path cost $c^*_j = c_j(m)$ such that $c_j(m) = \min_{n\in\{0,\ldots,k_i\}} c_j(n)$. Therefore, to guarantee optimality, this minimal cost to any matching goal has to take the place of the shortest path cost from the original ICTS root. Intuitively, the sum of individual costs of this root is highly optimistic, particularly if the teams are large. This is because often, agents will not move to their closest matching goal in an optimal solution, even without considering conflicts (see \ref{pathology}).

@@

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.40\textwidth}
			\centering
			\includegraphics[width=\linewidth]{img/counter-example-ep-2}
			\caption{Instance}
			\label{fig:counter-example:a}
		\end{subfigure}
		\begin{subfigure}{0.49\textwidth}
			\centering
			\includegraphics[width=\linewidth]{img/counter-example-tree}
			\caption{Pruned ICT that is traversed}
			\label{fig:counter-example:b}
		\end{subfigure}
		\caption{Counter-example for enhanced pruning combined with child pruning}
		\label{fig:counter-example}
	\end{figure}
	\begin{proof}
		Consider Figure \ref{fig:counter-example:a} with a red agent $r$, green agent $g$ and black agent $b$. It can be seen that $(3,8,5)$ with SIC $16$ is an optimal solution for this instance, which corresponds to $r$ moving aside, allowing $g$ to pass.  Say enhanced pruning with pairs is applied to these three agents at $(c^*_r,c^*_g,c^*_b) = (0,8,5)$. Let the enhanced pruning order be $(r,g)$, $(r,b)$, $(g,b)$ without loss of generality. $(r,g)$ is solvable but will leave for $g$ out of the two paths of length $8$ only the clock-wise path. $(r,b)$ will not change either MDD in this node. When $(g,b)$ is evaluated, it is found that the remaining clockwise path for $g$ and $b$'s path are incompatible (edge conflict) so $(0,9,5)$ and $(0,8,6)$ are generated. By repeating this order of enhanced pruning with child pruning, the ICT shown in \ref{fig:counter-example:b} is traversed. Notice that none of the nodes with SIC 16 in this ICT is a solution, meaning that enhanced pruning combined with child pruning will return a suboptimal solution to the instance in \ref{fig:counter-example:a}, making this pruning strategy suboptimal.
	\end{proof}

@@

	CBM solving all problems with one team (Figure \ref{fig:r-25-1-p} and \ref{fig:r-75-1-p}), problems equivalent to anonymous MAPF instances, is at first surprising but follows directly from the definition of the CBM method. CBM solves single teams as anonymous MAPF instances in polynomial time, meaning that solving a single team reduces to a polynomial-time procedure, namely SSP. Following the results reported in \cite{ma2016}, it is expected that as the number of teams approaches $k$, CBM performance will deteriorate and thus become more similar to the results for exhaustive methods here reported.
	%	Still, some implementations might be more optimized than others so this might be a factor in performance characteristics and in particular in relative performance differences.
	\section{Conclusions and Future work}
	\label{conclusions}
	\label{future}
	The goal of this research was to seek alternative approaches to solving MAPFM based on ICTS. In this, the research has succeeded, but the scalability of the algorithms leaves something to be desired. Exhaustive methods, like those discussed in detail in Subsection \ref{exhaustive}, cannot be expected to scale well as teams grow larger due to their complexity, no matter what optimizations are put in place. Nonetheless, as was shown in Section \ref{experiments}, the more advanced exhaustive ICTS algorithms consistently outperform the ICTS-m based algorithms. Compared to the SIC-adaptation of the CBM method \cite{baauw2021}, it was seen that exhaustive methods generally perform relatively poorly if there are fewer teams and fewer obstacles, while being competitive once teams grow smaller and maps become more constrained. Delineating under which conditions exhaustive methods outperform CBM and vice versa is something for further research.
